High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-18 doc/asplos2013/interface.tex
Working on the interface section

>>>>

The Legion Low-Level Runtime (LLR) interface organizes
functionality into objects.  Objects are grouped into
four different categories: machine objects for describing
the underlying architecture, events for capturing dependences,
deferred locks for performing synchronization, and physical regions
for managing data.  In all cases, the objects are light-weight
{\em handles} that can be passed around by value.  Furthermore
every handle is valid everywhere in the system.  This property allows 
a programmer to pass handles by value between computations without
having to reason about the distributed nature of the system.
We describe how we support this property in more detail in 
Section~\ref{sec:impl}.

\lstset{
  captionpos=b,
  language=C++,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  %literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  %morekeywords={region,coloring,partition,spawn,disjoint,aliased},
  %deletekeywords=float,
}
\subsection{Machine Interface}
\label{subsec:machmodel}
Listing~\ref{lst:machineapi} shows the machine interface.  The machine interface 
contains three different object types: {\tt Processor} (line 1), {\tt Memory} (line 17), and a 
singleton {\tt Machine} object (line 22).  For every processor (i.e. CPU core or GPU) in 
the system there exists a unique {\tt Processor} handle naming that processor.  
{\tt Processor} objects support a single {\tt spawn} operation (line 13-14) that will
launch a {\em task} on that processor.  Note that because the {\tt spawn} operation
is invoked on a processor handle, it is possible to launch a task on any
processor from anywhere in the system.  The {\tt spawn} call takes as arguments
the {\tt TaskFuncID} of the task to be run, a pointer to and size of arguments to
be passed to the task, and an optional {\tt Event} (described in Section~\ref{subsec:events})
that must trigger before the task can begin.  The result of the {\tt spawn} call
is an {\tt Event} that will trigger when the task completes.

For every memory in the system there also exists a unique {\tt Memory} handle. 
Different memories often, but not always, imply a different address space.  For
example, in a large cluster each node would have a different {\tt Memory} handle
corresponding to each node's memory.
However, if the cluster supported multiple GPUs then there would be a different
{\tt Memory} handle for each GPU's {\em zero-copy} memory which shares part of the 
address space with the corresponding node memory.  Memories will be used for describing
the placement of data described in Section~\ref{subsec:phyreg}.

The {\tt Machine} object is a singleton object that serves two purposes.  First, it
is the mechanism for initializing the runtime; at the start of a program the programmer
creates a {\tt Machine} object and invokes the {\tt run} method (line 30-31).
Second, the {\tt Machine} object provides an interface for the programmer to 
discover properties of the underlying machine (lines 35-43).  The {\tt Machine} object maintains
sets of all the {\tt Processor} and all the {\tt Memory} handles in the system.
The {\tt Machine} also maintains two relations between these sets:

\begin{itemize}
\item Processor-Memory: for each {\tt Processor} and each {\tt Memory} whether
the {\tt Processor} can directly access (read and write) the {\tt Memory} and 
latency and bandwidth properties if access is possible
\item Memory-Memory: for every pair of {\tt Memory} whether a copy can be
performed between the two memories and latency 
bandwidth properties if a copy is possible
\end{itemize}

The programmer can query these two relations by the {\tt get\_*\_affinity} calls
(lines 40-43).  These calls populate a vector with affinity structures (not shown)
that contain latency and bandwidth information.  The programmer can restrict the query
by specifying a specific processor or memory, otherwise information is returned
about all processors and memories in the system.

\begin{lstlisting}[float={t},label={lst:machineapi},caption={Machine Interface.}]
class Processor {
public:
  const unsigned id;
  typedef unsigned TaskFuncID;
  typedef void (*TaskFuncPtr)
            (const void *args,size_t arglen,Processor p);
  typedef map<TaskFuncID, TaskFuncPtr> TaskIDTable;

  enum Kind {
    CPU_PROC,GPU_PROC,// ... future processor types
  };

  Event spawn(TaskFuncID func_id,const void *args,size_t arglen,
              Event wait_on = Event::NO_EVENT) const;
};

class Memory {
public:
  const unsigned id;
};

class Machine {
public:
  Machine(int *argc, char ***argv,
          const Processor::TaskIDTable &task_table);
  enum RunStyle {
    ONE_TASK_ONLY,ONE_TASK_PER_NODE,
    ONE_TASK_PER_PROCESSOR,
  };
  void run(Processor::TaskFuncID task_id = 0, 
        RunStyle = ONE_TASK_ONLY, const void *args, size_t arglen);
public:
  static Machine* get_machine(void); // pointer to Machine
  // References to all processors and memories
  const set<Memory>& get_all_memories(void) const;
  const set<Processor>& get_all_processors(void) const;
  Processor::Kind get_processor_kind(Processor p) const;
  size_t get_memory_size(Memory m) const;
  // Query properties of underlying hardware
  int get_proc_mem_affinity(vector<ProcMemAffinity &result,
          Processor restrict_proc = 0, Memory restrict_mem = 0);
  int get_mem_mem_affinity(vector<MemMemAffinity> &result,
          Memory restrict_mem1 = 0, Memory restrict_mem2 = 0);
};
\end{lstlisting}

\subsection{Events}
\label{subsec:events}
Listing~\ref{lst:eventapi} shows the interface for events.  
{\tt Event} is the base type of events.  An instance of an {\tt Event} type 
names a unique event in the system.  {\tt NO\_EVENT} (line 5) is a special instance
of an event that by definition has always triggered.  The {\tt Event} interface
also supports testing whether an event has triggered (line 7) and waiting on
an event to trigger (line 9).  While these methods can be useful, the most common
use of events is passing them as preconditions for operations such as the {\tt spawn}
call described in Section~\ref{subsec:machmodel}.  The programmer can use the
{\tt merge\_events} call (line 11) to create an event corresponding to the 
conjunction of a set of events when multiple events are a precondition for an operation.

In most cases events are created by the runtime as the result of other
operations and the runtime is responsible for triggering these events.  Users 
can create a special type of event called a {\tt UserEvent} (line 14) that the
user has the power to trigger.  {\tt UserEvent} is a sub-type of {\tt Event} 
that has {\tt trigger} method (line 17).  Users can also create another
type of event called a {\tt Barrier} that will require multiple arrivals before
the event is triggered.  The user can manage the number of arrivals that must
be seen as well as how many arrivals occur at a time.  Note that since {\tt UserEvent}
and {\tt Barrier} are sub-types of {\tt Event} they can be used wherever an
{\tt Event} is required.

\begin{lstlisting}[float={t},label={lst:eventapi},caption={Event Interface.}]
class Event {
public:
  const unsigned id;
  const unsigned gen;
  static const Event NO_EVENT;
  // check if an event has triggered
  bool has_triggered() const;
  // wait on the event
  void wait() const;
  // Merge events together to a new event
  static Event merge_events(const set<Event> &to_merge);
};

class UserEvent : public Event {
public:
  static UserEvent create_user_event();
  void trigger() const;
};

class Barrier : public Event {
public:
  static Barrier create_barrier(unsigned expected_arrivals);
  void alter_arrival_count(int delta) const;
  void arrive(unsigned count = 1) const;
};
\end{lstlisting}

\subsection{Deferred Locks}
\label{subsec:locks}
Deferred locks are a new synchronization mechanism that has different semantics from
blocking locks.  From here on we refer to locks and deferred locks interchangeably, 
but will refer to blocking locks explicitly.  Listing~\ref{lst:lockapi} shows the 
interface for locks.  Unlike blocking locks, the {\tt lock} method (line 4) doesn't
block but instead always returns immediately with an event that will be triggered
when the lock has been acquired.  If the lock can be acquired immediately then a {\tt NO\_EVENT}
is returned.  The {\tt lock} method also takes an optional event
parameter indicating that the lock acquire shouldn't be performed until the event
has triggered.  Similarly, the {\tt unlock} method (line 5) takes an optional event parameter indicating
an event to wait for before performing the unlock operation.

Deferred locks provide a super-set of the functionality of blocking locks.  As their
name suggests, deferred locks can use the event corresponding to their lock acquire
operation to defer execution until the lock acquire has been granted.  Deferred
locks can also be converted back into a blocking lock by immediately waiting on the event
returned from a call to {\tt lock}.  In addition, deffered locks also provide {\tt mode} and
and {\tt exclusive} parameters that allow the user to specify whether other requests can acquire
the lock simultaneously.  A lock can only be acquired in one mode at a time.  The
{\tt exclusive} parameter specifies whether other owners are permitted once the
current lock request is granted.

An important difference between deferred locks and blocking locks is that the processor
that requests the lock doesn't have to be the one that uses it.  A common
convention in writing code with deferred locks is to acquire locks for a task being
launched.  Listing~\ref{lst:lockapi} illustrates this with a simple example.  The
{\tt launcher\_task} is calling a sub-task that is going to require the {\tt needed}
lock.  A lock request is issued and the resulting {\tt lock\_event} is used as
a precondition for launching the {\tt SUB\_TASK} task.  The unlock operation is then
precondition on the event corresponding to {\tt SUB\_TASK} task's completion.  The
sub-task can run safely while holding the lock and the lock is released when the task
sub-task completes.  Using deferred locks in this style minimizes the time
compute resources spend waiting on locks by ensuring tasks only run when their needed
locks have already been acquired.

\begin{lstlisting}[float={t},label={lst:lockapi},caption={Deferred Lock Interface and Example.}]
class Lock {
public:
  const unsigned id;
  Event lock(unsigned mode = 0, bool exclusive, 
              Event wait_on = Event::NO_EVENT) const;
  void unlock(Event wait_on = Event::NO_EVENT) const;
  // Create a new lock, destroy an existing lock
  static Lock create_lock();
  void destroy_lock();
};

void launcher_task(const void *args, size_t arglen, Processor p) {
  ...
  // Unpack lock from arguments
  Lock needed = ...
  // Acquire lock
  Event lock_event = needed.lock();
  // Launch task
  Event task_event = p.spawn(SUB_TASK,NULL,0,lock_event);
  // Release lock
  needed.unlock(task_event);
  ...
}
\end{lstlisting}

\subsection{Physical Regions}
\label{subsec:phyreg}
